---
title: "Cyclistic Data Analysis Notebook: Analyze Phase"
author: "Donnie Minnick"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
# output: rmarkdown::html_vignette
# vignette: >
#   %\VignetteIndexEntry{prepare}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
---

## Context

This notebook documents part of a data analysis I completed in the capstone experience for the Google Data Analytics Professional Certificate program on Coursera.

You can find the Github repository for my work [here](https://github.com/dtminnick/cyclistic).

## Analyze Phase

This phase involves formatting and transforming data, identifying patterns, drawing conclusions and making data-driven decisions.

### Hypothesis

To complete this analysis, I need to setup a comparison between casual riders and members, exploring whether their usage patterns differ and, if so, how they differ.  I believe there will be two fundamental usage patterns: some customers will be inclined towards work commutes during the week, while others will be more inclined toward recreational use during weekends.

**Null Hypothesis**: There is no significant difference in the usage patterns between casual riders and members of the bike-sharing service. Both groups are equally likely to use the service for work commutes during the week and recreation on weekends.

**Alternative Hypothesis**: There is a significant difference in the usage patterns between casual riders and members of the bike-sharing service. The usage patterns may vary, with one group being more inclined towards work commutes during the week, while the other group is more inclined towards recreational use on weekends.

This hypothesis statement sets up the comparison between casual riders and members.  I will test this hypothesis and assess whether there is enough evidence to reject the null hypothesis in favor of the alternative hypothesis.

The outcome of this analysis will then inform strategies Cyclistic can use to increase profits by converting casual riders to members.

### Load Libraries

I am using libraries from the tidyverse collection of R packages and supplementing with the knitr, RColorBrewer and sf libraries for table formatting, color palettes and manipulation of spatial data, i.e. to generate maps.

```{r load_libraries, message = FALSE}
packages <- c("dplyr", "ggplot2", "knitr", "lubridate", "RColorBrewer", "sf", "stringr", "tidyr")

installed_packages <- packages %in% rownames(installed.packages())

if(any(installed_packages == FALSE)) {
  
  install.packages(packages[!installed_packages])
  
}

invisible(lapply(packages, library, character.only = TRUE))
```

### Load Trip Data

Load the trip sample data.

```{r load_trip_data}
load("C:/R/Packages/cyclistic/data/trip.rda")
```

The data frame contains a random selection of `r format(nrow(trip), big.mark = ",", scientific = FALSE)` trips from the 2022 data frame I created during the prepare and process phases of the project.

### Color Palette

Specify a color-blind-friendly palette of eight colors for use with charts and visuals.

```{r specify_palette}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

## Exploratory Analysis

### Trips by Member Type

Establish the proportion of trips taken by casual riders and members.

```{r trips_by_member_type}
trips_by_member_type <- trip %>%
  group_by(member_type) %>%
  summarise(trip_count = n()) %>%
  mutate(percent = round(trip_count / sum(trip_count), 2)) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                       across(where(is.character), ~ "Combined")))

kable(trips_by_member_type,
      col.names = c("Member Type", "Trip Count", "Percent"),
      caption = "Trips by Member Type",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r"))
```

So 59% of trips are taken by members and 41% are taken by casual riders, i.e. four in every ten customers are casual riders.  This is our conversion opportunity.  Plot these percentages.

```{r plot_trips_by_member_type}
ggplot(trips_by_member_type[1:2,], aes(x = member_type, y = trip_count)) +
  geom_bar(stat = "identity", fill = cbPalette[2]) + 
  labs(title = "Trips by Member Type",
       subtitle = "Four in every ten customers are casual riders; this is our conversion opportunity.",
       x = "Member Type",
       y = "Trip Count") + 
  geom_text(aes(label = paste(percent * 100, "%", sep = "")), colour = "white", vjust = 1.5) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

### Observation 1 - Casual Rider Conversion Opportunity

Given the data sample, I estimate that casual riders contribute 40% of trips to Cyclistic's bike-sharing activity.

### Trips by Trip Type

Trips can be point-to-point (i.e. starts at one station and ends at another) or loop (i.e. starts and ends at the same station).  Show trips by trip type.

```{r trips_by_trip_type}
trips_by_trip_type <- trip %>%
  group_by(member_type, trip_type) %>%
  tally() %>%
  spread(member_type, n) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                       across(where(is.character), ~ "Combined"))) %>%
  mutate(total = rowSums(across(where(is.numeric))))

kable(trips_by_trip_type,
      col.names = c("Trip Type", "Casual", "Member", "Total"),
      caption = "Trips Crosstab: Trip Counts by Trip Type & Member Type",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r"))
```

Most trips are point-to-point; 5% are loops.

Whether riders are using bike-sharing for recreational purposes or commuting, it's unlikely they would retain the rental beyond the leg of a given trip as this would require them to paid rental fees for idle time.  So while there may be fewer loop-type trips, the data is not captured in a manner that this dynamic can be better quantified.  

To better quantify this, we would need rider-specific information to match trips by rider.  Assuming we then find multiple trips with common start and end stations, we could tag these as a trip type of loop.

### Trips by Ride Type

Do customers show any preference in ride type, i.e. classic bicycle versus electric?  Determine the percent of rides taken using each ride type offered.

```{r trips_by_ride_type}
trips_by_ride_type <- trip %>%
  group_by(ride_type) %>%
  summarise(trip_count = n()) %>%
  mutate(percent = round(trip_count / sum(trip_count), 2)) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                     across(where(is.character), ~ "Combined")))

kable(trips_by_ride_type,
      col.names = c("Ride Type", "Trip Count", "Percent"),
      caption = "Trips by Ride Type",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r"))
```

A significant majority of customers, approximately 60%, prefer classic bikes, while a substantial portion, around 37%, opt for electric bikes. This distribution suggests a notable preference for traditional bikes among the customer base, with a significant but comparatively smaller interest in electric alternatives.  Plot the percentages.

```{r plot_trips_by_ride_type}
ggplot(trips_by_ride_type[1:3,], aes(x = ride_type, y = trip_count)) +
  geom_bar(stat = "identity", fill = cbPalette[2]) + 
  labs(title = "Trips by Ride Type",
       subtitle = "Customers show preference for classic bicycles, with significant interest in electric alternatives.",
       x = "Ride Type",
       y = "Trip Count") + 
  geom_text(aes(label = paste(percent * 100, "%", sep = "")), colour = "white", vjust = 1.5) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

How does this preference vary when comparing customer types?

```{r trips_by_member_and_ride_type_v1}
trips_by_member_and_ride_type_v1 <- trip %>%
  group_by(member_type, ride_type) %>%
  summarise(trip_count = n()) %>%
  pivot_wider(names_from = member_type, values_from = trip_count) %>%
  replace(is.na(.), 0) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                   across(where(is.character), ~ "Combined"))) %>%
  mutate(total = rowSums(across(where(is.numeric))))

kable(trips_by_member_and_ride_type_v1,
      col.names = c("Ride Type", "Casual", "Member", "Total"),
      caption = "Trips by Member & Ride Type",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r"))
```

51% of casual riders used classic bicycles, compared to 66% of members.  And 41% of casual riders prefer electric bicycles, compared to 66% of members.  Plot the percentages.

```{r plot_trips_by_member_and_ride_type_v1}
trips_by_member_and_ride_type_v2 <- trip %>%
  group_by(member_type, ride_type) %>%
  summarise(trip_count = n()) %>%
  mutate(percent = round(trip_count / sum(trip_count), 2))

ggplot(trips_by_member_and_ride_type_v2, aes(x = ride_type, y = trip_count)) +
  geom_bar(stat = "identity", fill = cbPalette[2]) + 
  labs(title = "Trips by Member Type & Ride Type",
     subtitle = "A greater proportion of casual riders prefer electric bicycles compared to members.",
     x = "Ride Type",
     y = "Trip Count") + 
  facet_grid(cols = vars(member_type)) + 
  geom_text(aes(label = paste(percent * 100, "%", sep = "")), colour = "white", vjust = 1.5) + 
  guides(fill=guide_legend(title="Ride Type")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

Based on these proportions, it appears that casual riders may have a greater preference for electric bicycles when compared to members.

Is there sufficient evidence to draw this conclusion, i.e. is the proportion of casual riders using electric bicycles statistically greater than the same proportion of members?

Perform a proportions test to confirm whether the observed proportion of electric bicycle riders among casual riders is greater than the proportion for members with 95% confidence.

```{r proportions_test_electric}
prop.test(x = c(41308, 50269), n = c(101726, 148274), alternative = "greater")
```

Given the test’s p-value of 2.2e-16, which is less than the alpha = 0.05 significance level, we may conclude that the proportion of customers with a preference for electric bicycles in the two groups is significantly different, i.e. casual riders have a greater preference for electric bicycles.

### Observation 2 - Electric Bicycle Preference

So Cyclistic should ensure that its fleet of electric bicycles is sufficiently large enough to meet the needs of casual riders that prefer them.  A limited supply of electric bicycles may hinder the company's ability to convert casual riders to member status.

### Notes on Docked Bicycles

Presumably, docking stations are special bicycle racks that lock the bicycles and only release them by computer control.  Perhaps Cyclistic uses both docked and dockless systems where riders must leave a bike at a designated station or can leave them anywhere within a geo-fenced area, e.g. within city limits, when they finish using them.  It's also not clear whether docked bicycles are electric or not.

In this sample, members have not used any docked bicycles, and casual rider use is 8% of total trips recorded for casual users.  Given this limited use, I will treated docked bicycles as classic bicycles.

### Trips by Month

Generate a time-based view of rides to show the number of trips made per month by each customer type.

```{r trips_by_member_type_and_month_v1}
trips_by_member_type_and_month_v1 <- trip %>%
  mutate(month = month(start_date_time, label = TRUE, abbr = TRUE)) %>%
  group_by(month, member_type) %>%
  summarise(trip_count = n()) %>%
  pivot_wider(names_from = month, values_from = trip_count) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                 across(where(is.character), ~ "Combined"))) %>%
  mutate(total = rowSums(across(where(is.numeric))))

kable(trips_by_member_type_and_month_v1,
      col.names = c("Member Type", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec", "Total"),
      caption = "Trips by Member Type & Month",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r"))
```

Peak volumes occur in the summer months of June, July and August when, presumably, the weather is most conducive to riding bicycles.  Alternatively, ride volume is at it's lowest point in the winter months of December, January and February when cold temperatures and winter weather makes it difficult to ride bicycles.  Plot this pattern.

```{r plot_trips_by_member_type_and_month_v2}
trips_by_member_type_and_month_v2 <- trip %>%
  mutate(month = month(start_date_time, label = TRUE, abbr = TRUE)) %>%
  group_by(month, member_type) %>%
  summarise(trip_count = n())

ggplot(trips_by_member_type_and_month_v2, aes(x = month, y = trip_count, fill = member_type)) +
  geom_bar(stat = "identity") + 
  labs(title = "Trips by Member Type & Month",
       subtitle = "The seasonal volume of trips peaks in summer for both customer types.",
       x = "Member Type",
       y = "Trip Count") + 
  guides(fill=guide_legend(title="Member Type")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

This seasonal pattern is essentially the same for both casual riders and members.

### Observation 3 - Leverage Seasonality in Marketing

Given the steady rise in trips ahead of the summer peak, the early spring months are a good time to campaign to casual riders to purchase memberships.  Beyond the practical benefits of bike-sharing for work commute, it's common to observe a natural rise in bike riding during the spring and summer months. Several factors contribute to this seasonal pattern:

**Weather**: Warmer temperatures and longer daylight hours in spring and summer create more favorable conditions for outdoor activities, including biking.

**Seasonal Events**: Many cities and communities organize biking events, races, or group rides during the warmer months, attracting more people to participate.

**School Vacations**: With schools often having breaks during the summer, families may engage in recreational activities together, including biking.

**Tourism**: In tourist destinations, the spring and summer months often see an increase in visitors who may explore the area on bikes.

**Health and Fitness Goals**: People may be more motivated to engage in physical activities, such as biking, as part of their health and fitness goals, and this motivation may peak in the warmer months.

### Trips by Duration

Generate a summary of the trip duration column.

```{r trip_duration_summary}
summary(trip$trip_duration_in_minutes)
```

Mean trip duration is approximately 15 minutes with significant variation around that average.  Assuming a rider can cover one mile in 3-5 minutes, the approximate distance traveled would be in the range of 3-6 miles on average per trip.  Plot the distribution of trips by member type.

```{r trips_by_member_type_and_duration}
ggplot(trip, aes(x = trip_duration_in_minutes, fill = member_type)) +
  geom_histogram(binwidth = .5, alpha =.5, position = "dodge") + 
  labs(title = "Trips by Member Type & Trip Duration",
     subtitle = "Members take more short-duration trips (less than 20 minutes) compared to casual riders.",
     x = "Trip Duration",
     y = "Trip Count") + 
  guides(fill=guide_legend(title="Member Type")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

This chart shows that members have a higher volume of trips with a duration of less than twenty minutes.  For trips with duration of twenty minutes or greater, casual riders and members have roughly the same volume.

What might account for this difference?  There may be a connection between purpose of the trips (i.e. work commute versus recreation) and duration.  The source data does not include data points for purpose, but I might be able to infer this by breaking down the data by day of week.

### Trips by Category of Day

Generate a plot that shows trips by member type and category of day (i.e. workweek versus weekend).

```{r trips_by_member_type_and_day_category}
ggplot(trip, aes(x = trip_duration_in_minutes, fill = category_of_day)) +
  geom_histogram(binwidth = .5, alpha =.5, position = "dodge") +
  facet_grid(rows = vars(member_type)) + 
  labs(title = "Trips by Member Type & Trip Duration",
   subtitle = "Most of the short-duration trips occur during the work week.",
   x = "Trip Duration",
   y = "Trip Count") + 
  guides(fill=guide_legend(title="Category of Day")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

So during weekdays, members are taking more trips than casual members and on weekends, the distribution of trips for both customer types is similar.

Are these trips constrained at all by geography or citiscape?  If someone is using the bike-sharing service for work commute, then this might be a consideration, i.e. if the citiscape doesn't provide bike lanes, it's likely to be safer and more convenient to use public transportation.  Someone commuting to work is likely to also be constrained by schedule, at least when it comes to arriving to work on time.

The dataset does not include information on why customers use Cyclistic's service, i.e. the purpose of their trip.

It does include information, the start and end date/time of each trip, we can use to infer purpose.  For example, a rider using  the service primarily to commute to and from work will have trips during the weekdays.  And riders using the service for other purposes, e.g. recreation, will likely have trips during weekend days.  

In addition, we might be able to infer purpose using geolocation data.  If, for example, there are a high volume of trips starting or ending at destinations in the city, e.g. parks, attractions, cultural and culinary locations, we might infer the service is being used for non-commute purposes.

### Trips by Day of Week

```{r trips_by_day_of_week}
trips_by_member_type_and_day_of_week_v1 <- trip %>%
  group_by(day_of_week, member_type) %>%
  summarise(trip_count = n()) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                      across(where(is.character), ~ "Combined"))) %>%
  pivot_wider(names_from = day_of_week, values_from = trip_count) %>%
  mutate(total = rowSums(across(where(is.numeric))))

kable(trips_by_member_type_and_day_of_week_v1,
      col.names = c("Member Type", "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Total"),
      caption = "Trips by Member Type & Day of Week",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r", "r", "r", "r"))
```

```{r plot_trips_by_day_of_week}
trips_by_member_type_and_day_of_week_v2 <- trip %>%
  group_by(day_of_week, member_type) %>%
  summarise(trip_count = n())

ggplot(trips_by_member_type_and_day_of_week_v2, aes(x = day_of_week, y = trip_count, fill = member_type)) +
  geom_bar(stat = "identity") + 
  labs(title = "Trips by Member Type & Day of Week",
       subtitle = "While overall trips peak on Saturdays; for members, trips peak in mid-week.",
       x = "Day of Week",
       y = "Trip Count") + 
  guides(fill = guide_legend(title = "Member Type")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

Saturday is the busiest day of the week for Cyclistic in terms of overall trip volume.

However, mid-week is the peak in trips for members, occuring on Tuesday, Wednesday and Thursday.

Although not definitive, this data does support the hypothesis that members are using bikes to commute to and from work during the week, and that casual riders are using Cyclistic for more recreational opportunities the city provides.

### Trips by Hour of Day

```{r trips_by_hour}
start_trips <- trip %>%
  group_by(start_hour_of_day, day_of_week, member_type) %>%
  summarise(trip_count = n()) %>%
  mutate(hour_type = "Start") %>%
  rename(hour_of_day = start_hour_of_day)

end_trips <- trip %>%
  group_by(end_hour_of_day, day_of_week, member_type) %>%
  summarise(trip_count = n()) %>%
  mutate(hour_type = "End") %>%
  rename(hour_of_day = end_hour_of_day)

trips_by_hour <- rbind(start_trips, end_trips)
```



```{r trips_by_start_hour}
trips_by_start_hour <- trips_by_hour %>%
  filter(hour_type == "Start") %>%
  select(-member_type) %>%
  group_by(hour_of_day, day_of_week) %>%
  summarise(trip_count = sum(trip_count)) %>%
  mutate(hour_of_day = str_pad(as.character(hour_of_day), 2, "left", "0")) %>%
  arrange(hour_of_day) %>%
  pivot_wider(names_from = day_of_week, values_from = trip_count) %>%
  ungroup() %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                      across(where(is.character), ~ "Combined"))) %>%
  mutate(total = rowSums(across(where(is.numeric))))

kable(trips_by_start_hour,
      col.names = c("Hour", "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Total"),
      caption = "Trips by Start Hour",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r", "r", "r", "r"))
```



```{r trips_by_end_hour}
trips_by_end_hour <- trips_by_hour %>%
  filter(hour_type == "Start") %>%
  select(-member_type) %>%
  group_by(hour_of_day, day_of_week) %>%
  summarise(trip_count = sum(trip_count)) %>%
  mutate(hour_of_day = str_pad(as.character(hour_of_day), 2, "left", "0")) %>%
  arrange(hour_of_day) %>%
  pivot_wider(names_from = day_of_week, values_from = trip_count) %>%
  ungroup() %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                      across(where(is.character), ~ "Combined"))) %>%
  mutate(total = rowSums(across(where(is.numeric))))

kable(trips_by_end_hour,
      col.names = c("Hour", "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Total"),
      caption = "Trips by Start Hour",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r", "r", "r", "r"))
```



```{r plot_trips_by_hour}
ggplot(trips_by_hour, aes(x = hour_of_day, y = trip_count, fill = hour_type)) +
  geom_bar(stat = "identity") + 
  labs(title = "Trips by Hour of Day",
   subtitle = "Trips peak between the hours of 7:00-9:00 am and 4:00-6:00 pm.",
   x = "Hour of Day",
   y = "Trip Count") + 
  guides(fill=guide_legend(title="Hour Type")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

As you might expect, trips tend to rise over the course of the morning hours, starting at roughly 7:00 am and peak in the mid-afternoon hours (i.e. 4:00-6:00 pm), then decline during the evening hours.  4:00-6:00 pm hours are the peak time for both starting and ending trips.  

Curious to know whether this pattern changes when considering weekdays versus weekend days.

There is a similar pattern across all days of week, except that weekend days (Saturday and Sunday) have a pattern that more closely resembles a normal distribution, with rides steadily increasing, beginning at 8:00 am, peaking at approximately 3:00 pm, then steadily declining in the late afternoon and evening hours.  This pattern is distinct from the weekday pattern described above.

Also, when the data is separated by weekday, there is another pattern in the weekday data: an early morning peak that occurs between the hours of 7:00-8:00 am.  

These patterns suggest that during the week, customers are using the bike-sharing service to commute to and from work, while on weekends, customers are using the service for other, presumably recreational purposes during the daytime hours.

Is there any additional insight to be gained by viewing this data by member type?

```{r plot_trips_by_hour_of_day}
start_trips <- trip %>%
  group_by(start_hour_of_day, day_of_week, member_type) %>%
  summarise(ride_count = n()) %>%
  mutate(type = "Start") %>%
  rename(hour_of_day = start_hour_of_day)

end_trips <- trip %>%
  group_by(end_hour_of_day, day_of_week, member_type) %>%
  summarise(ride_count = n()) %>%
  mutate(type = "End") %>%
  rename(hour_of_day = end_hour_of_day)

trips_by_hour_v2 <- rbind(start_trips, end_trips) %>%
  select(-type)

ggplot(trips_by_hour, aes(x = hour_of_day, y = ride_count, fill = member_type)) +
  geom_bar(stat = "identity") +
  facet_grid(rows = vars(day_of_week)) + 
  labs(title = "Trips by Member Type & Day of Week",
    subtitle = "....",
    x = "Hour of Day",
    y = "Trip Count") + 
  guides(fill=guide_legend(title = "Member Type")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

The same patterns persist; however, there are casual members who are using the bike-sharing service for commutes during the week.  In addition, there are a greater number of casual riders who are using the service on weekend days.  

These patterns suggest a two-part strategy for converting casual riders to members:

* Highlighting the benefits of member status for commutes to and from work, and
* Engaging these casual riders as members for recreational rides as well.

To support this conversion, the marketing department should highlight the potential expense savings of a membership for a typical commuter, and the added benefits when the service is used for recreation in addition to commuting.

### Common Daily Trips

What are the most common daily trips, and how do these trips vary by day of week and member type?

```{r}
most_common_starts <- trip %>%
  group_by(day_of_week, start_station_name) %>%
  summarise(trip_count = n()) %>%
  arrange(desc(trip_count)) %>%
  mutate(percent = round(trip_count / sum(trip_count), 3)) %>%
  top_n(n = 30)
  
most_common_ends <- trip %>%
  group_by(day_of_week, end_station_name) %>%
  summarise(trip_count = n()) %>%
  arrange(desc(trip_count)) %>%
  mutate(percent = round(trip_count / sum(trip_count), 3)) %>%
  top_n(n = 30)

```

### Chicago Data Portal

The [Chicago Data Portal](https://data.cityofchicago.org/) lets you find city data and create maps and graphs about the city.  Users can freely download the data for their analyses.

I downloaded shape files from the portal to create maps of Cyclistic bike-sharing stations, visualize station usage, and understand the relationship between stations and features of the city, i.e. communities, parks, and landmarks.

I downloaded the corresponding shape files and use them to generate maps.

```{r read_shape_files}
communities <- st_read("C:/R/Packages/cyclistic/inst/extdata/chicago/communities/geo_export_03f8c85a-4634-44fb-bcb9-4653575ff11b.shp")

parks <- st_read("C:/R/Packages/cyclistic/inst/extdata/chicago/parks/geo_export_cc5e1418-25db-49e7-a6fd-af8782efc515.shp")

landmarks <- st_read("C:/R/Packages/cyclistic/inst/extdata/chicago/landmarks/Chicago_Landmarks_June2012.shp")

national_landmarks <- st_read("C:/R/Packages/cyclistic/inst/extdata/chicago/national/LandmarksNationalRegister_nov2012.shp")

routes <- st_read("C:/R/Packages/cyclistic/inst/extdata/chicago/routes/geo_export_a6173042-ca11-4e91-bb92-fa19586fbb96.shp")
  
```

```{r}
start_stations <- trip %>%
  group_by(category_of_day, start_station_name, start_longitude, start_latitude) %>%
  summarise(trip_count = n()) %>%
  distinct_all() %>%
  arrange(desc(trip_count)) %>%
  ungroup() %>%
  filter(category_of_day == "Weekend") %>%
  filter(between(row_number(), 1, 100)) %>%
  mutate(type = "Start") %>%
  rename(station_name = start_station_name, 
         longitude = start_longitude, 
         latitude = start_latitude)
  
end_stations <- trip %>%
  group_by(category_of_day, end_station_name, end_longitude, end_latitude) %>%
  summarise(trip_count = n()) %>%
  distinct_all() %>%
  arrange(desc(trip_count)) %>%
  ungroup() %>%
  filter(category_of_day == "Weekend") %>%
  filter(between(row_number(), 1, 100)) %>%
  mutate(type = "End") %>%
  rename(station_name = end_station_name, 
         longitude = end_longitude, 
         latitude = end_latitude)

stations <- rbind(start_stations, end_stations)

ggplot(data = communities) +
    geom_sf(fill = "grey", alpha = 0.3) +
    geom_point(data = stations, aes(x = longitude, 
                                    y = latitude, 
                                    size = trip_count,
                                    color = trip_count,
                                    alpha=I(0.1))) +
    coord_sf(xlim = c(-87.85, -87.52), 
             ylim = c(41.65, 42.07), 
             expand = TRUE) +
  facet_grid(cols = vars(type)) +
  labs(title = "Stations by Trip Count",
    subtitle = "....",
    x = "Longitude",
    y = "Latitude") + 
  guides(fill=guide_legend(title = "Trip Count")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```
```{r}
start_stations <- trip %>%
  group_by(category_of_day, start_station_name, start_longitude, start_latitude) %>%
  summarise(trip_count = n()) %>%
  distinct_all() %>%
  arrange(desc(trip_count)) %>%
  ungroup() %>%
  filter(category_of_day == "Workweek") %>%
  filter(between(row_number(), 1, 100)) %>%
  mutate(type = "Start") %>%
  rename(station_name = start_station_name, 
         longitude = start_longitude, 
         latitude = start_latitude)
  
end_stations <- trip %>%
  group_by(category_of_day, end_station_name, end_longitude, end_latitude) %>%
  summarise(trip_count = n()) %>%
  distinct_all() %>%
  arrange(desc(trip_count)) %>%
  ungroup() %>%
  filter(category_of_day == "Workweek") %>%
  filter(between(row_number(), 1, 100)) %>%
  mutate(type = "End") %>%
  rename(station_name = end_station_name, 
         longitude = end_longitude, 
         latitude = end_latitude)

stations <- rbind(start_stations, end_stations)

ggplot(data = communities) +
    geom_sf(fill = "grey", alpha = 0.3) +
    geom_point(data = stations, aes(x = longitude, 
                                    y = latitude, 
                                    size = trip_count,
                                    color = trip_count,
                                    alpha=I(0.1))) +
    coord_sf(xlim = c(-87.85, -87.52), 
             ylim = c(41.65, 42.07), 
             expand = TRUE) +
  facet_grid(cols = vars(type)) +
  labs(title = "Stations by Trip Count",
    subtitle = "....",
    x = "Longitude",
    y = "Latitude") + 
  guides(fill=guide_legend(title = "Trip Count")) + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

### Parks

```{r}
ggplot() +
  geom_sf(data = communities, fill = "white") + 
  geom_sf(data = parks, fill = "#E69F00") + 
  labs(title = "Trips by Member Type & Day of Week",
    subtitle = "....",
    x = "Longitude",
    y = "Latitude") + 
  scale_fill_manual(values = cbPalette) + 
  theme_light()
```

### Landmarks

```{r}
ggplot() +
  geom_sf(data = communities, fill = "white") + 
  geom_sf(data = landmarks, fill = "blue")
```

### National Landmarks

```{r}
ggplot() +
  geom_sf(data = communities, fill = "white") + 
  geom_sf(data = national_landmarks, fill = "blue")
```

### Bike Routes

```{r}
ggplot() +
  geom_sf(data = communities, fill = "white") + 
  geom_sf(data = routes, fill = "blue")
```



