---
title: "Cyclistic Data Analysis Notebook: Process Phase"
author: "Donnie Minnick"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
# output: rmarkdown::html_vignette
# vignette: >
#   %\VignetteIndexEntry{prepare}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
---

## Context

This notebook documents part of a data analysis I completed in the capstone experience for the Google Data Analytics Professional Certificate program on Coursera.

You can find the Github repository for my work [here](https://github.com/dtminnick/cyclistic).

## Process Phase

This phase involves cleaning and transforming data, maintaining data integrity, testing data and verifying and reporting on cleaning results.

I am using libraries from the tidyverse collection of R packages and supplementing with the knitr library to format tables contained within this notebook.

```{r load_libraries, message = FALSE}
packages <- c("dplyr", "ggplot2", "knitr", "lubridate", "stringr", "tidyr")

installed_packages <- packages %in% rownames(installed.packages())

if(any(installed_packages == FALSE)) {
  
  install.packages(packages[!installed_packages])
  
}

invisible(lapply(packages, library, character.only = TRUE))
```

The code chunk below loads the source data gathered in the prepare phase of this project.

```{r load_source_data}
load("C:/R/Packages/cyclistic/data/source.rda")

trip <- source
```

The data frame contains `r format(ncol(source), big.mark = ",", scientific = FALSE)` columns and `r format(nrow(source), big.mark = ",", scientific = FALSE)` rows of data.

### Missing Values

I'll start by checking for missing values.

The code below counts the number of missing values by column and presents the counts in a long-form table.

```{r missing_values}
missing_values <- trip %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  pivot_longer(cols = everything()) %>%
  rename(column_name = name,
         row_count = value)

kable(missing_values,
      col.names = c("Column Name", "Row Count"),
      caption = "Missing Values by Column",
      format.args = list(big.mark = ","),
      align = c("l", "r"))
```

There are rows with missing location data, including station name, id and location coordinates.  Since the missing data can't be imputed, I will omit these rows from the data frame.

```{r drop_nas}
trip <- trip %>%
  na.omit()
```

Dropping these rows reduces the overall number of rows in the source data to `r format(nrow(source), big.mark = ",", scientific = FALSE)` rows for analysis.

### Data Transformations

To facilitate additional cleaning, will format columns in the source data frame and ensure there are no duplicate rows in the data frame.

The code chunk below makes the following updates to the data frame:
* Renames columns,
* Formats latitude and longitude to number with five positions following the decimal,
* Formats start and end date/time as Date objects,
* Creates a trip duration variable by dividing the difference between the start and end date times by 60,
* Creates a trip type variable that is coded as 'loop' if the start and end stations are the same or 'point-to-point' if they are not,
* Creates variables to capture the start and end hour for each trip, and
* Recodes ride type (Classic, Docked, Electric) and member type (Casual, Member), and
* Creates a day of week variable based on the the start date/time.

```{r format_data_frame}
trip <- trip %>%
  rename(ride_type = rideable_type,
         start_date_time = started_at,
         end_date_time = ended_at,
         start_latitude = start_lat,
         start_longitude = start_lng,
         end_latitude = end_lat,
         end_longitude = end_lng,
         member_type = member_casual) %>%
  mutate(start_latitude = round(start_latitude, 5),
         start_longitude = round(start_longitude, 5),
         end_latitude = round(end_latitude, 5),
         end_longitude = round(end_longitude, 5),
         start_date_time = lubridate::ymd_hms(start_date_time),
         end_date_time = lubridate::ymd_hms(end_date_time),
         trip_duration_in_minutes = round(as.numeric((end_date_time - start_date_time) / 60), 2),
         trip_type = if_else(start_station_name == end_station_name, "Loop", "Point-to-Point"),
         start_hour_of_day = lubridate::hour(start_date_time),
         end_hour_of_day = lubridate::hour(end_date_time),
         ride_type = case_when(ride_type == "classic_bike" ~ "Classic",
                               ride_type == "docked_bike" ~ "Docked",
                               ride_type == "electric_bike" ~ "Electric"),
         member_type = case_when(member_type == "casual" ~ "Casual",
                                 member_type == "member" ~ "Member"),
         day_of_week = lubridate::wday(start_date_time, TRUE, TRUE))
```

### Same-Day Trips

I'm going to filter the data frame to include only those rows that are same-day trips since that's the most common scenario in this data.

```{r filter_by_dates}
trip <- trip %>%
  filter(str_sub(start_date_time, 1, 10) == str_sub(end_date_time, 1, 10))
```

### Trip Duration

There are some rows with a trip duration of zero or negative values.  There are also rows with durations of close to 24 hours.  I will filter the data frame to include trips with a duration in the range of 5-960 minutes, i.e. eight hours.

```{r filter_by_duration}
trip <- trip %>%
  filter(trip_duration_in_minutes >= 5 & trip_duration_in_minutes <= 960)
```

This leaves a range of duration values from `r min(source$trip_duration_in_minutes)` to `r max(source$trip_duration_in_minutes)` minutes.

### Station Names

Some station names are causing duplicates.  The following code addresses them by cleaning the station names.

```{r clean_station_names}
trip <- trip %>%
  mutate(start_station_name = str_replace(start_station_name, "Public Rack - ", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, "Public Rack - ", "")) %>%
  mutate(start_station_name = str_replace(start_station_name, "Public  Rack - ", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, "Public  Rack - ", "")) %>%
  mutate(start_station_name = str_replace(start_station_name, "Pubic Rack - ", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, "Pubic Rack - ", "")) %>%
  mutate(start_station_name = str_replace(start_station_name, "City Rack - ", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, "City Rack - ", "")) %>%
  mutate(start_station_name = str_replace(start_station_name, " - midblock", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, " - midblock", "")) %>%
  mutate(start_station_name = str_replace(start_station_name, " - midblock south", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, " - midblock south", "")) %>%
  mutate(start_station_name = str_replace(start_station_name, " - Charging", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, " - Charging", "")) %>%
  mutate(start_station_name = str_replace(start_station_name, " (Temp)", "")) %>%
  mutate(end_station_name = str_replace(end_station_name, " (Temp)", ""))
```

And the following code addresses variants of the same station names.

```{r update_station_names}
trip <- trip %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "N Shore Channel Trail & Argyle Ave", 
                                          "N Shore Channel Trail & Argyle St")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "N Shore Channel Trail & Argyle Ave", 
                                        "N Shore Channel Trail & Argyle St")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Ewing Ave & 106th St", 
                                          "Ewing Ave & 106th St NW")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Ewing Ave & 106th St", 
                                        "Ewing Ave & 106th St NW")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Ewing Ave & 96th St", 
                                          "Ewing Ave & 96th St N")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Ewing Ave & 96th St", 
                                        "Ewing Ave & 96th St N")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Mt Greenwood Library - North", 
                                          "Mt Greenwood Library N")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Mt Greenwood Library - North", 
                                        "Mt Greenwood Library N")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Kedzie & 103rd St - West", 
                                          "Kedzie & 103rd St - W")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Kedzie & 103rd St - West", 
                                        "Kedzie & 103rd St - W")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Lafayette Ave & 87th St SW", 
                                          "Lafayette Ave & 87th St - SW")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Lafayette Ave & 87th St SW", 
                                        "Lafayette Ave & 87th St - SW")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "63rd & Western Ave - south corner", 
                                          "63rd & Western Ave S")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "63rd & Western Ave - south corner", 
                                        "63rd & Western Ave S")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "63rd & Western Ave - north corner", 
                                          "63rd & Western Ave N")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "63rd & Western Ave - north corner", 
                                        "63rd & Western Ave N")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Kedzie Ave & 61st Pl", 
                                          "Kedzie Ave & 61st Pl E")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Kedzie Ave & 61st Pl", 
                                        "Kedzie Ave & 61st Pl E")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Prairie Ave & 47th St ", 
                                          "Prairie Ave & 47th St")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Prairie Ave & 47th St ", 
                                        "Prairie Ave & 47th St")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Woodlawn Ave & 63rd St - SE", 
                                          "Woodlawn Ave & 63rd St S")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Woodlawn Ave & 63rd St - SE", 
                                        "Woodlawn Ave & 63rd St S")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Woodlawn Ave & 63rd St - NE", 
                                          "Woodlawn Ave & 63rd St N")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Woodlawn Ave & 63rd St - NE", 
                                        "Woodlawn Ave & 63rd St N")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Prairie Ave & Garfield Blvd", 
                                          "Prairie Ave & Garfield Blvd N")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Prairie Ave & Garfield Blvd", 
                                        "Prairie Ave & Garfield Blvd N")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Kedvale Ave & 63rd St", 
                                          "Kedvale Ave & 63rd St W")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Kedvale Ave & 63rd St", 
                                        "Kedvale Ave & 63rd St W")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Kedzie Ave &amp; 62nd Pl", 
                                          "Kedzie Ave & 62nd Pl")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Kedzie Ave &amp; 62nd Pl", 
                                        "Kedzie Ave & 62nd Pl")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Kedzie Ave & 61st Pl", 
                                          "Kedzie Ave & 61st Pl W")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Kedzie Ave & 61st Pl", 
                                        "Kedzie Ave & 61st Pl W")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Bissell St & Armitage Ave*", 
                                          "Bissell St & Armitage Ave")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Bissell St & Armitage Ave*", 
                                        "Bissell St & Armitage Ave")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "WestChi", 
                                          "WEST CHI-WATSON")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "WestChi", 
                                        "WEST CHI-WATSON")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Base - 2132 W Hubbard", 
                                          "Base - 2132 W Hubbard Warehouse")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Base - 2132 W Hubbard", 
                                        "Base - 2132 W Hubbard Warehouse")) %>%
  mutate(start_station_name = str_replace(start_station_name, 
                                          "Base - 2132 W Hubbard", 
                                          "Base - 2132 W Hubbard Warehouse")) %>%
  mutate(end_station_name = str_replace(end_station_name, 
                                        "Base - 2132 W Hubbard", 
                                        "Base - 2132 W Hubbard Warehouse"))
```

Ensure data frame has distinct rows.

```{r remove_duplicate_rows}
trip <- trip %>% 
  distinct_all()
```

```{r save_trip}
write.csv(trip, 
          "../inst/extdata/output/trip.csv",
          row.names = FALSE)

usethis::use_data(trip, overwrite = TRUE)
```

The code below ensures the source data frame has distinct rows and saves the cleaned data frame.  It now contains `r ncol(source)` columns and `r nrow(source)` rows of data.

### Calendar

Create a calendar table.

```{r create_calendar}
start_date <- as.Date("2022-01-01")
end_date <- as.Date("2022-12-31")

calendar <- data.frame(date = seq.Date(start_date, end_date, by = "1 day"))

calendar <- calendar %>%
  mutate(year = lubridate::year(date),
         quarter = lubridate::quarter(date),
         month = lubridate::month(date),
         week = lubridate::week(date),
         day_of_month = lubridate::day(date),
         year_and_quarter = paste(lubridate::year(date), "-", str_pad(lubridate::quarter(date), 2, "left", "0"), sep = ""),
         year_and_month = paste(lubridate::year(date), "-", str_pad(lubridate::month(date), 2, "left", "0"), sep = ""),
         month_long_name = lubridate::month(date, TRUE, FALSE),
         month_short_name = lubridate::month(date, TRUE, TRUE),
         day = lubridate::wday(date),
         day_long_name = lubridate::wday(date, TRUE, FALSE),
         day_short_name = lubridate::wday(date, TRUE, TRUE))

write.csv(calendar, 
          "../inst/extdata/output/calendar.csv",
          row.names = FALSE)

usethis::use_data(calendar, overwrite = TRUE)
```

### Destination

Create a destination table.

Purpose is to compare trips to popular destinations in the city.

```{r create_destination}
destination <- read.delim("../inst/extdata/source/destination.txt",
                          header = TRUE,
                          sep = ";",
                          na.strings = c(""),
                          stringsAsFactors = FALSE)

write.csv(destination, 
          "../inst/extdata/output/destination.csv",
          row.names = FALSE)

usethis::use_data(destination, overwrite = TRUE)
```

### Data Integrity

Regarding data integrity, I noted the following considerations:

* The data is generally reliable, even if there are some missing values for location-related information, 
* The data is original, first-party data collected by Cyclistic, 
* The data is comprehensive enough for purposes of this analysis, and
* The data is current, with data sets that cover the period January 2013 through November 2023.

## Next Steps

The next step in this project is to analyze the data in preparation for sharing.
